---
title: "Scratch Insplico Analysis"
project:
  type: website
  output-dir: docs
format: 
  html:
    code-fold: show
    toc: true
    toc-location: right
    theme: Sketchy
number-sections: true
execute: 
  echo: true
  cache: true
  warning: false
---

```{r setup}
library(dplyr)
library(data.table)
library(stringr)
library(tidyr)
library(ggplot2)
library(betareg)
library(see)

```

## Material

Insplico uses a tab delimited file where all exons are listed and which are going to be quantified. [Here](https://raw.githubusercontent.com/liniguez/InsplicoAnalysis/main/exons_subset_insplico.tab) you can get a subset of one used to run insplico. Insplico is ran for each sample individually and then the output of all the runs can be **merged** into a single big data. [Here](https://raw.githubusercontent.com/liniguez/InsplicoAnalysis/main/subset_final_insplico.tab) you can get a subset of a run. Here I analyzed two single cell libraries NB292 and NB293. I got the clustering of the cells based on gene expression. Alignment file was then split per sample into different clusters, based on cell barcodes and used as input for Insplico.

::: callout-note
It is important to mention that each row form Insplico output correspond to each row from tab delimited exon file. More info about insplico in [here](https://gitlab.com/aghr/insplico)
:::

## Read tables

```{r read}
exons<-fread("exons_subset_insplico.tab", nrows = 1000) %>% mutate(ROW=row_number())
insp<-fread("subset_final_insplico.tab", nrows = 1000) %>%
  mutate(ROW=row_number())
```

Columns UPINT_MINLEN,DOWNINT_MINLEN,INTLEN_CONSIDERED,EX_MINLEN,EX_MAXLEN are equal for all samples and can be added to exons variable.

```{r}
exons<-insp %>% select(UPINT_MINLEN,DOWNINT_MINLEN,INTLEN_CONSIDERED,EX_MINLEN,EX_MAXLEN) %>%
  cbind(exons)
```

For differential splicing analysis (DSA) we will only use 4 columns:

-   PSI_EX

-   EXON_INCLSPLICED_2UPSTRM

-   EXON_INCLSPLICED_2DOSTRM

-   EXON_EXCL

```{r}
insp<-insp %>% select(-c(UPINT_MINLEN,DOWNINT_MINLEN,INTLEN_CONSIDERED,EX_MINLEN,EX_MAXLEN)) %>%
  select(ROW,matches(c("PSI_EX","EXON_INCLSPLICED_2UPSTRM","EXON_INCLSPLICED_2DOSTRM","EXON_EXCL"))) %>%
  select(!matches("WITHIRREADS")) %>% 
  pivot_longer(-ROW,names_to = "temp", values_to = "values",values_drop_na = TRUE)
```

Modify column name for getting the sample and cluster values

```{r}
samples.clusters<-stringr::str_split(insp$temp,pattern = "_") %>%
  lapply(function(x){paste0(rev(x)[1],".",rev(x)[2])}) %>% unlist
clusters<-gsub(".+\\.",replacement = "",samples.clusters,perl=T)
samples<-gsub("\\..+",replacement = "",samples.clusters,perl=T)
```

Classify the value into one of the 4 columns selected.

```{r}
features<-case_when(grepl("PSI_EX",insp$temp) ~ "PSI_EX",
                   grepl("EXON_INCLSPLICED_2UPSTRM",insp$temp) ~ "EXON_INCLSPLICED_2UPSTRM",
                   grepl("EXON_INCLSPLICED_2DOSTRM",insp$temp) ~ "EXON_INCLSPLICED_2DOSTRM",
                   grepl("EXON_EXCL",insp$temp) ~ "EXON_EXCL",
                   .default = NA)
```

Add sample and count feature to the tibble

```{r}
insp<-insp %>% mutate("sample.cluster"=samples.clusters,"sample"=samples,"cluster"=clusters,"feature"=features)

```

## Filter Exons

We then need to get the count of supporting reads for inclusion/exclusion and for that we'll do; for long reads:

> (EXON_INCLSPLICED_2UPSTRM+EXON_INCLSPLICED_2DOSTRM)/2 + EXON_EXCL

for short reads it would be something like:

> EXON_INCLSPLICED_2UPSTRM +EXON_INCLSPLICED_2DOSTRM + EXON_EXCL

Here we are using the mentioned values for assigning a quality score to each PSI based on vast-tools quality parameters ([more info](https://github.com/vastgroup/vast-tools)).

```{r}

covScore <- function(total_reads, exon_excl, exon_inclspliced_2upstrm, exon_inclspliced_2dostrm){
  case_when(
    total_reads >= 100 & (exon_excl >= 20 | ( (exon_inclspliced_2upstrm >= 20 & exon_inclspliced_2dostrm >= 15) | (exon_inclspliced_2upstrm >= 15 & exon_inclspliced_2dostrm >= 20) )) ~ "SOK",
    total_reads < 100 & (exon_excl >= 20 | ( (exon_inclspliced_2upstrm >= 20 & exon_inclspliced_2dostrm >= 15) | (exon_inclspliced_2upstrm >= 15 & exon_inclspliced_2dostrm >= 20) )) ~ "OK",
    exon_excl >= 15 | ( (exon_inclspliced_2upstrm >= 15 & exon_inclspliced_2dostrm >= 10) | (exon_inclspliced_2upstrm >= 10 & exon_inclspliced_2dostrm >= 15) ) ~ "LOW",
    exon_excl >= 10 | ( (exon_inclspliced_2upstrm >= 10 & exon_inclspliced_2dostrm >= 5) | (exon_inclspliced_2upstrm >= 5 & exon_inclspliced_2dostrm >= 10) ) ~ "VLOW",
    TRUE ~ "N"
  )
}

rawinsp<-insp %>% filter(!is.na(feature)) %>%
  pivot_wider(id_cols = c(sample,sample.cluster,cluster,ROW),names_from = "feature", values_from = "values") %>% 
  mutate(
    PSI = round(100 * PSI_EX, 2),
    TOT_READS = EXON_INCLSPLICED_2UPSTRM + EXON_INCLSPLICED_2DOSTRM + EXON_EXCL,
    INC = round(TOT_READS * PSI_EX, 1),
    EXC = round(TOT_READS * (1 - PSI_EX), 1),
    COV = covScore(total_reads=TOT_READS, 
                   exon_excl=EXON_EXCL, 
                   exon_inclspliced_2upstrm=EXON_INCLSPLICED_2UPSTRM, 
                   exon_inclspliced_2dostrm=EXON_INCLSPLICED_2DOSTRM),
    Q = paste0(COV, ",", 
               TOT_READS, ",", 
               EXON_INCLSPLICED_2UPSTRM, ",", 
               EXON_INCLSPLICED_2DOSTRM, ",", 
               EXON_EXCL, "@", 
               INC, ",", 
               EXC)
  )  %>% dplyr::select(sample,sample.cluster,cluster,ROW,ROW, PSI, Q)

```

Usually for DSA exons with Quality of *N* are ignored.

Additionally, constitutive exons (PSI \> 90) and cryptic ones ( PSI \<10) are also removed.

Here we have only two samples one per group, but more complex filters can be applied (this one is very basic, latter other filters are shown).


## No Replicates

Lets work only with the PSI values from cluster 1. The question is whether exons are differentially spliced between the two samples of cluster 1. In this case we have only two samples, and we will compare both. First let's filter the cryptic and constitutuve events.


```{r}
psiFilt<-rawinsp %>% filter(cluster==1)%>%
  mutate(PSI_fil=ifelse(grepl("^N",Q,perl=T),NA,PSI)) %>%
  filter(!is.na(PSI_fil))%>%
  pivot_wider(id_cols = c(ROW),
              names_from = "sample",values_from = "PSI_fil",values_fill = NA)
psiFilt_noCryp_noConst<-psiFilt%>%   
  mutate(prop_const=apply(select(.,-c(ROW)),1,function(x){sum(x>90)/length(x)}),
         prop_cryp=apply(select(.,-c(ROW)),1,function(x){sum(x<10)/length(x)}))%>%
  filter(!is.na(prop_const) & prop_const<=0.5 & prop_cryp<=0.5)
```
This is how the events to be analyzed look like:

```{r, fig.width=4, fig.height=4}
psiFilt_noCryp_noConst %>% ggplot(aes(x=NB292, y=NB293))+geom_point(size=0.5)+theme_bw()
```

For this part I used similar methods as the ones reported in the paper of [betAS](https://rnajournal.cshlp.org/content/early/2024/01/26/rna.079764.123). This methods are still under discussion and comments for improvement are more than welcome. The needed table can be found in [here](https://raw.githubusercontent.com/liniguez/InsplicoAnalysis/main/table_betAS.txt).

```{r}
betasmat<-read.table("table_betAS.txt") #This table is provided by the betAS package.
givenCovfindIncr <- function(cov, maxDevRef){
  if(cov > max(maxDevRef[,1])){
    incr <- 1
  }else{
    p <- which(cov == maxDevRef$cov)
    incr <- maxDevRef$maxIncr[p]
  }
  return(incr)
}

#I applied a recursive function with a limit of 10^7 sim for a pseudo p-value aproximation.
getpseudop<-function(inc1,excl1,inc2,excl2, seed=12032024){
  nsim=1000
  p_zero=1
  set.seed(seed)
  while(p_zero==1 & nsim<10^7){ 
    r1<-rbeta(nsim,inc1,excl1)
    r2<-rbeta(nsim,inc2,excl2)
    p_zero<-sum((r1-r2>0))/nsim
    nsim<-nsim*10
  }
  psudopval<-1-abs((p_zero-0.5)*2)
}

```

We need again the number of reads per sample that support inclusion/exclusion of the exon. We'll need to modify a bit betAS pipeline. Pseudo-pvalue will be adjusted with fdr.

```{r}
getINC<-function(x){
  y<-gsub(pattern="@",replacement = ",",x) %>% strsplit(split=",") %>% unlist()
  inc<-as.numeric(y[3])+as.numeric(y[4])
  return(inc)
}
getEXC<-function(x){
  y<-gsub(pattern="@",replacement = ",",x) %>% strsplit(split=",") %>% unlist()
  exc<-as.numeric(y[5])
  return(exc)
}

resSimBeta_temp<-rawinsp %>% filter(cluster==1 & ROW %in% psiFilt_noCryp_noConst$ROW)%>%
  rowwise()%>%
  mutate(INC=getINC(Q),
         EXCL=getEXC(Q)*2,  # EXON_EXCL*2
         INC=ifelse(INC==0,givenCovfindIncr(round(EXCL),betasmat),INC),
         EXCL=ifelse(EXCL==0,givenCovfindIncr(round(INC),betasmat),EXCL)) %>%
  ungroup() %>%
  select(ROW,sample,cluster,INC,EXCL) %>%
  pivot_wider(id_cols = c("ROW","cluster"),names_from = "sample",values_from = c("INC","EXCL"))

resBETAS<-sapply(1:length(resSimBeta_temp$ROW),function(x){
  res<-getpseudop(inc1=as.numeric(resSimBeta_temp[x,"INC_NB292"]),
                  excl1=as.numeric(resSimBeta_temp[x,"EXCL_NB292"]),
                  inc2=as.numeric(resSimBeta_temp[x,"INC_NB293"]),
                  excl2=as.numeric(resSimBeta_temp[x,"EXCL_NB293"]))
  return(res)
})
resSimBeta_temp$psudopval<-resBETAS
resSimBeta<-resSimBeta_temp %>% group_by(cluster) %>% mutate(sudopadj= p.adjust(psudopval,method = "fdr"))

```

This is the filtered result:

```{r}
exons_results<-psiFilt_noCryp_noConst %>% mutate(dPSI=NB293-NB292) %>%
  left_join(resSimBeta, by=c("ROW")) %>%
  mutate(DiffSP=ifelse(sudopadj<0.1 & abs(dPSI)>15,"DS","ns"),DiffSP=factor(DiffSP,levels=c("ns","DS")) ) 

exons_results %>% filter(DiffSP=="DS") %>% left_join(exons, by="ROW") %>% arrange(sudopadj) %>% data.table()

```

```{r, fig.height=4,fig.width=6}
data4plot<-exons_results %>% filter(DiffSP=="DS") %>% left_join(exons, by="ROW") %>%
  arrange(sudopadj)%>%
  apply(1,function(x){
  nsim<-1000
  inc1<-as.numeric(x[8])
  excl1<-as.numeric(x[10])
  inc2<-as.numeric(x[9])
  excl2<-as.numeric(x[11])
  r1<-rbeta(nsim,inc1,excl1)
  r2<-rbeta(nsim,inc2,excl2)
  tibble("a"=r1, "b"=r2) %>%
    mutate(ROW=x[1],EXON_ID=x[29], GENE=x[23], C1=x[15],STARTS=x[16],ENDS=x[17],C2=x[18] )
}) %>% do.call(rbind,.)
data4plot %>%
  mutate(GENE=toupper(GENE)) %>%
  rowwise() %>% 
  mutate(GENEEXON=paste0(GENE,".",EXON_ID))%>%
  ungroup()%>%
  mutate(GENEEXON=forcats::fct_inorder(GENEEXON))%>%
  ggplot(aes(x=GENEEXON))+
  see::geom_violinhalf(aes(y=b, fill="b"), color="gray30",trim = T,scale="width")+
  see::geom_violinhalf(aes(y=a, fill="a"),flip=TRUE, color="gray30",trim=T,scale="width")+
  theme_bw(10)+
  scale_fill_manual(values=c("b"="orange","a"="royalblue"), name="Samples")+
  labs(y="simulated PSI", x="Gene.ExonID")+
  theme(axis.text.x = element_text(angle = 30,hjust = 1))
```

## With Replicates

Now lets assume we have replicates and the different clusters are replicates of the sample. Here again we need to filter constitutive and cryptic exons. First, we'll check what is the proportion of clusters from each sample where the number of reads supporting the inclusion/exclusion is greater than 15, and filter if the proportion is greater than 30%. This is too lax, but it is because of the toy data used, usually this proportion should be above 50%.

```{r}
insp_noN<-rawinsp %>% 
  mutate(PSI_fil=ifelse(grepl("^N",Q,perl=T),NA,PSI)) %>%
  filter(!is.na(PSI_fil))

nsamples_pergroup<-rawinsp  %>% select(sample,sample.cluster) %>% unique() %>% group_by(sample) %>% reframe(N_tot=n())

summaryEvpergr<-insp_noN %>%
  group_by(sample,ROW) %>% reframe(N=n(), meanPSI=mean(PSI),
                                      sdPSI=sd(PSI), minPSI=min(PSI), maxPSI=max(PSI),
                                      medianPSI=median(PSI)) %>%
  left_join(nsamples_pergroup, by="sample") %>% mutate(Prop=N/N_tot)

#Check for events in at least 30% of the samples in here in more than 4 clusters

evin2check<- summaryEvpergr %>% 
    filter(Prop>0.3) %>% group_by(ROW) %>% reframe(okgrps=n()) %>% filter(okgrps>1)
tempSum<-summaryEvpergr %>%  filter(ROW %in% unique(evin2check$ROW)) 


```

We then check what is the proportion of clusters in each sample where an exon is consitituve or cryptic. We filtered out exons where in more than 75% it showed a pattern of constitiuve or cryptic in both samples.

```{r}

EvNOTCons<-insp_noN %>% 
    filter( ROW %in% unique(evin2check$ROW)) %>%
    filter(PSI >90 ) %>% 
    group_by(ROW,sample) %>% reframe(N_c=n()) %>% full_join(tempSum, by=c("ROW","sample")) %>% 
    mutate(N_c=ifelse(is.na(N_c),0,N_c),
           Prop_cons=N_c/N) %>% 
    pivot_wider(id_cols = ROW,names_from = sample,values_from = Prop_cons,values_fill = 0) %>%
    rename(all_of(c(Test="NB292",Ref="NB293")))  %>% filter(Test < 0.5 | Ref < 0.5)
  
EvNOTCry<-insp_noN %>% 
    filter( ROW %in% unique(evin2check$ROW)) %>%
    filter(PSI <10 ) %>% 
    group_by(ROW,sample) %>% reframe(N_c=n()) %>% full_join(tempSum, by=c("ROW","sample")) %>% 
    mutate(N_c=ifelse(is.na(N_c),0,N_c),
           Prop_cons=N_c/N) %>% 
    pivot_wider(id_cols = ROW,names_from = sample,values_from = Prop_cons,values_fill = 0) %>%
    rename(all_of(c(Test="NB292",Ref="NB293")))  %>% filter(Test < 0.5 | Ref < 0.5)
  
evin2check<-EvNOTCons  %>%
    full_join(EvNOTCry, by=c("ROW"), suffix = c("_constitutive","_cryptic")) %>%
    filter((!is.na(Ref_cryptic) & !is.na(Test_cryptic))&
             (!is.na(Ref_constitutive)& !is.na(Test_constitutive)))%>%
    select(ROW) %>% unique() 
```

For assesing DSA with replicates I used beta regression on the samples and p-values were adjusted with fdr

```{r}
ref="NB292"
test="NB293"
Rep_res<-insp_noN %>% select(ROW,sample, PSI) %>%
    right_join(evin2check, by=c("ROW")) %>% 
    filter(sample %in% c(ref,test)) %>%
    split(.$ROW) %>%
    lapply(function(x){
      evrow=x$ROW %>% unique()
      temp<-x%>%
        mutate(TestFactor=factor(sample,levels=c(ref,test))) %>%
        filter(!is.na(TestFactor))
      if(length(unique(temp$TestFactor))>1){
        pvalor<-temp %>%
          mutate(PSI_EX=PSI/100,PSI_EX=ifelse(PSI_EX==0,0.01,
                                              ifelse(PSI_EX==1,0.99,PSI_EX))) %>%
          betareg(PSI_EX ~TestFactor|1,data = .) %>% summary() %>%
          .$coefficients %>% .$mean
        pvalor<-pvalor[paste0("TestFactor",test),'Pr(>|z|)']
      }else{pvalor <-1}
      means<-temp %>% group_by(TestFactor) %>% reframe(M=mean(PSI))
      ref_means<-means %>% filter(TestFactor==ref)
      test_means<-means  %>% filter(TestFactor!=ref)
      dPSI<-test_means %>%
        mutate(dpsi=M-ref_means$M, name=paste0("TestFactor",TestFactor)) %>%
        select(name,dpsi)
      result_out<-tibble("ROW"=unique(x$ROW),test_PSI=test_means$M,
                         ref_PSI=ref_means$M, "dPSI"=dPSI$dpsi,"pvalue"=pvalor)
      return(result_out)
    }) %>% do.call(rbind,.)
Rep_res$padj<-p.adjust(Rep_res$pvalue, method = "fdr")

subset(Rep_res,padj<0.1 & abs(dPSI)>15) %>% data.table()
```

```{r}
data2plot<- insp_noN %>% filter(ROW %in% subset(Rep_res,padj<0.1 & abs(dPSI)>15)$ROW)

data2plot %>% left_join(exons, by="ROW") %>% rowwise() %>% mutate(NAME=paste0(toupper(GNAME),".",EXON_ID)) %>%  
ggplot(aes(x=sample, y=PSI))+
  geom_boxplot(outlier.shape = NA)+
  geom_point(aes(color=as.factor(cluster)), position = position_dodge(width=0.5))+
  ylim(c(0,100))+
  facet_wrap(~NAME)+
  labs(y="PSI")+
  theme_bw()+
  scale_color_viridis_d()+
  theme(axis.title.x = element_blank(), legend.position = "none")

```
